%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Adjectives describe Nouns}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{\textit{ni}}
% \index{\textit{ike}}
% \index{\textit{jaki}}
% \index{\textit{lawa}}         
\index{\textit{len}}
% \index{\textit{lili}}
% \index{\textit{mute}}
\index{\textit{nasa}}
\index{\textit{seli}}
% \index{\textit{sewi}}
% \index{\textit{tomo}}
% \index{\textit{utala}}
\index{complicated}
\index{evil}
\index{bad}
\index{trash}
\index{nasty}
\index{dirty}
\index{head}
\index{leading}
\index{main}
\index{clothing}
\index{little}
\index{a lot}
\index{many}
\index{weird}
\index{silly}
\index{stupid}
\index{crazy}
\index{drunk}
\index{strange}
\index{hot}
\index{warm}
\index{sky}
\index{superior}
\index{high}
\index{building}
\index{house}
\index{construction}
\index{fight}
\index{battle}
\index{war}
\index{burn}
\subsection*{Vocabulary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{supertabular}{p{2,5cm}|ll}
% \textbf{\dots ike} && \textit{adjective}: bad, negative, wrong, evil, overly complex \\ % no-dictionary
% \textbf{\dots ike} && \textit{adverb}: bad, negative, wrong, evil, overly complex \\ % no-dictionary
% \textbf{ike} && \textit{noun}: negativity, badness, evil \\ % no-dictionary
% \textbf{ike} && \textit{verb intransitive}: to be bad, to suck \\ % no-dictionary
% \textbf{ike (e \dots)} && \textit{verb transitive}: to make bad, to worsen \\ % no-dictionary
%  && \\ % no-dictionary
% \textbf{\dots jaki} && \textit{adjective}: dirty, gross, filthy, obscene \\ % no-dictionary
% \textbf{\dots jaki} && \textit{adverb}: dirty, gross, filthy \\ % no-dictionary
% \textbf{jaki} && \textit{noun}: dirt, pollution, garbage, filth, feces \\ % no-dictionary
% \textbf{jaki (e \dots)} && \textit{verb transitive}: to pollute, to dirty \\ % no-dictionary
%  && \\ % no-dictionary
% \textbf{\dots lawa} && \textit{adjective}: main, leading, in charge \\ % no-dictionary
% \textbf{\dots lawa} && \textit{adverb}: main, leading, in charge \\ % no-dictionary
% \textbf{lawa} && \textit{noun}: head, mind \\ % no-dictionary
% \textbf{lawa (e \dots)} && \textit{verb transitive}: to lead, to control, to rule, to steer \\ % no-dictionary
%  && \\ % no-dictionary
\textbf{\dots len} && \textit{adjective}: dressed, clothed, costumed, dressed up \\ % no-dictionary
\textbf{len} && \textit{noun}: clothing, cloth, fabric, network, internet \\ % no-dictionary
\textbf{len (e \dots)} && \textit{verb transitive}: to wear, to be dressed, to dress \\ % no-dictionary
 && \\ % no-dictionary
% \textbf{\dots lili} && \textit{adjective}: small, little, young, a bit, short, few, less \\ % no-dictionary
% \textbf{\dots lili} && \textit{adverb}: small, little, young, a bit, short, few, less \\ % no-dictionary
% \textbf{lili} && \textit{noun}: smallness, youth, immaturity \\ % no-dictionary
% \textbf{lili (e \dots)} && \textit{verb transitive}: to reduce, to shorten, to shrink, to lessen \\ % no-dictionary
%  && \\ % no-dictionary
% \textbf{\dots mute} && \textit{adjective}: many, very, much, several, a lot, abundant, numerous, more \\ % no-dictionary
% \textbf{\dots mute} && \textit{adverb}: many, very, much, several, a lot, abundant, numerous, more \\ % no-dictionary
% \textbf{mute} && \textit{noun}: amount, quantity \\ % no-dictionary
% \textbf{mute (e \dots)} && \textit{verb transitive}: to make many or much \\ % no-dictionary
%  && \\ % no-dictionary
\textbf{\dots nasa} && \textit{adjective}: silly, crazy, foolish, drunk, strange, stupid, weird \\ % no-dictionary
\textbf{\dots nasa} && \textit{adverb}: silly, crazy, foolish, drunk, strange, stupid, weird \\ % no-dictionary
\textbf{nasa} && \textit{noun}: stupidity, foolishness, silliness, nonsense, idiocy, obtuseness, muddler \\ % no-dictionary
\textbf{nasa (e \dots)} && \textit{verb transitive}: to drive crazy, to make weird \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots ni} && \textit{adjective demonstrative pronoun}: this, that \\ % no-dictionary
\textbf{ni} && \textit{noun demonstrative pronoun}: this, that \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots seli} && \textit{adjective}: hot, warm, cooked \\ % no-dictionary
\textbf{\dots seli} && \textit{adverb}: hot, warm, cooked \\ % no-dictionary
\textbf{seli} && \textit{noun}: fire, warmth, heat \\ % no-dictionary
\textbf{seli (e \dots)} && \textit{verb transitive}: to heat, to warm up, to cook \\ % no-dictionary
%  && \\ % no-dictionary
% \textbf{\dots sewi} && \textit{adjective}: superior, elevated, religious, formal \\ % no-dictionary
% \textbf{\dots sewi} && \textit{adverb}: superior, elevated, religious, formal \\ % no-dictionary
% \textbf{sewi} && \textit{noun}: high, up, above, top, over, on \\ % no-dictionary
% \textbf{sewi} && \textit{verb intransitive}: to get up \\ % no-dictionary
% \textbf{sewi (e \dots)} && \textit{verb transitive}: to lift \\ % no-dictionary
%  && \\ % no-dictionary
% \textbf{\dots tomo} && \textit{adjective}: urban, domestic, household \\ % no-dictionary
% \textbf{\dots tomo} && \textit{adverb}: urban, domestic, household \\ % no-dictionary
% \textbf{tomo} && \textit{noun}: indoor constructed space, e.g. house, home, room, building \\ % no-dictionary
% \textbf{tomo (e \dots)} && \textit{verb transitive}: to build, to construct, to engineer \\ % no-dictionary
%  && \\ % no-dictionary
% \textbf{\dots utala} && \textit{adjective}: fighting \\ % no-dictionary
% \textbf{\dots utala} && \textit{adverb}: fighting \\ % no-dictionary
% \textbf{utala} && \textit{noun}: conflict, disharmony, fight, war, battle, attack, violence \\ % no-dictionary
% \textbf{utala (e \dots)} && \textit{verb transitive}: to hit, to strike, to attack, to compete against \\ % no-dictionary
\end{supertabular} \\
% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{adjective}
\index{noun!compound}
\subsection*{Adjectives}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We had already learned about adjectives as predicate adjectives as a part of a predicate phrase. 
Adjectives are comparable with adverbs but in Toki Pona some more complex.

Toki Pona has a very minimal vocabulary. 
Many words do not exist in this language. 
Therfore, we often have to combine various words together.
Adjectives are used to describe nouns. 
The noun \textit{jan} is described here with the adjective \textit{pona}.
As you can see the adjective goes after the main noun. 
This will undoubtedly seem incredibly awkward to you if you only speak English. 
However, many languages do this. 
Adjective slots are possible after nouns and, as already described, on \textit{li} as predicate adjective.

\begin{supertabular}{p{5,5cm}|ll}
jan pona && friend (good person) \\
\end{supertabular} 

For example, there is no word that means 'friend'. 

\index{victim}
\index{fork}
\index{spoon}
\index{knife}
\begin{supertabular}{p{5,5cm}|ll}
jan pakala && an injured person, victim, etc. \\
ilo moku && an eating utensil (fork/spoon/knife) \\
\end{supertabular} 

\index{adjective!more than one}
You can add until three adjective to a noun. 
One adjective should not be used more than once.

\index{soldier}
\begin{supertabular}{p{5,5cm}|ll}
jan utala && soldier  \\
jan utala nasa && stupid soldier  \\
jan utala nasa mute && many stupid soldiers  \\
jan utala nasa ni && this stupid soldier  \\
\end{supertabular} 

\index{adjective!more than one!\textit{ni}}
\index{adjective!more than one!\textit{mute}}
As you might have noticed, \textit{ni} and \textit{mute} as adjectives come at the end of the phrase. 
The reason for this is that the phrases build as you go along, so the adjectives must be put into an organized, logical order. 
Notice the differences in these two phrases.

\index{sidekick}
\begin{supertabular}{p{5,5cm}|ll}
jan utala nasa && stupid soldier  \\
jan nasa utala && fighting fool \\
\end{supertabular}

\index{adjective!combination}
Here are some handy adjective combinations using words that you've already learned and that are fairly common.

\index{ugly}
\index{enemy}
\index{leader}
\index{child}
\index{god}
\index{adult}
\index{lover}
\index{prostitute}
\index{mud}
\index{swamp}
\index{city}
\index{town}
\index{we}
\index{us}
\index{they}
\index{them}
\index{pretty}
\index{attractive}
\index{alcohol}
\index{beer}
\index{wine}
\index{restroom}
\index{pretty}
\index{chat room}
\index{conference room}
\begin{supertabular}{p{5,5cm}|ll}
ike lukin && ugly  \\
pona lukin && pretty, attractive \\
jan ni li pona lukin && That person is pretty. \\
jan ike && enemy \\
jan lawa && leader \\
jan lili && child \\
jan sewi && saint, God, Flying Spaghetti Monster \\
jan suli && adult \\
jan unpa && lover, prostitute \\
ma telo && mud, swamp \\
ma tomo && city, town \\
mi mute && we, us \\
ona mute && they, them \\
telo nasa && alcohol, beer, wine \\
tomo telo && restroom \\
ilo suno && flashlight \\ 
\end{supertabular} 
 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{possessive pronouns}
\index{pronoun}
\index{possessive pronouns!\textit{mi}}
\index{possessive pronouns!\textit{sina}}
\subsection*{Possessive Pronouns}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A possessive pronoun expresses a property or affiliation and is placed after the corresponding (composite) noun. 
This means that for a noun with adjectives, the possessive pronoun is placed after the adjectives. 
For a noun without adjectives, the possessive pronoun is located after the noun.
In these examples are \textit{mi}, \textit{sina} and \textit{ona} possessive pronouns. 

\begin{supertabular}{p{5,5cm}|ll}
tomo pona mi && my nice house \\
ma sina && your country \\
telo ona && his/her/its water \\
\end{supertabular} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{demonstrative pronoun}
\index{pronoun!demonstrative}
\subsection*{The Demonstrative Pronoun \textit{ni}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

The demonstrative pronoun is a kind of word with which the speaker refers to an  item of conversation. 
The demonstrative pronoun \textit{ni} can be used both like an adjective and like a noun.  
A slot for an adjective demonstrative pronoun is therefore possible after a noun. 

\begin{supertabular}{p{5,5cm}|ll}
jan ni li pona. && This bloke is good. \\
jan li lukin e ijo ni. && The guy's looking at this thing. \\
\end{supertabular}

A noun demonstrative pronoun is used instead of the noun. 
Slots for noun demonstrative pronouns therefore correspond to the positions of noun slots in the sentence. 

\begin{supertabular}{p{5,5cm}|ll}
ni li pona... && This is good. \\
jan li lukin e ni. && The guy looks at that one. \\
\end{supertabular}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
\subsection*{Practice (Answers: Page~\pageref{'adjectives'})}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


Please write down your answers and check them afterwards. 


\begin{supertabular}{p{5,5cm}|ll}



	
\end{supertabular}

See how well you can read the following poem. 

\begin{supertabular}{p{5,5cm}|ll}
mi jo e kili. && \\ % no-dictionary
ona li ' pona li ' lili. && \\ % no-dictionary
mi moku lili e kili lili. && \\ % no-dictionary
\end{supertabular} 

Try to translate these sentences. 
You can use the tool \textit{Toki Pona Parser} (\cite{www:rowa:02}) for spelling and grammar check. 

\begin{supertabular}{p{5,5cm}|ll}
The leader drank dirty water. &&   \\ % no-dictionary
I need a fork.   &&   \\ % no-dictionary
An enemy is attacking them.   &&   \\ % no-dictionary
That bad person has strange clothes.   &&  \\  % no-dictionary
We drank a lot of vodka.   &&   \\ % no-dictionary
Children watch adults.   &&   \\ % no-dictionary
 && \\ % no-dictionary
mi lukin e ni. &&  \\ % no-dictionary
mi lukin sewi e tomo suli.  &&    \\ % no-dictionary
seli suno li seli e tomo mi.  &&   \\ % no-dictionary
jan lili li wile e telo kili.  &&  \\ % no-dictionary
ona mute li nasa e jan suli. * &&  \\ % no-dictionary
\end{supertabular} 

* Notice how even though \textit{nasa} is typically an adjective, it is used as a verb here. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% eof
