%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Direct Objects}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vocabulary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{supertabular}{p{2,5cm}|ll}
%
\index{e}
\textbf{\dots e \dots} && \textit{separator}: An 'e' introduces a direct object. \\ && Don't use 'e' before or after the other separators. \\ % no-dictionary
 && \\ % no-dictionary
%
\index{ijo}
\textbf{\dots ijo} && \textit{adjective}: of something \\ % no-dictionary
\textbf{\dots ijo} && \textit{adverb}: of something \\ % no-dictionary
\textbf{ijo} && \textit{noun}: thing, something, stuff, anything, object \\ % no-dictionary
\textbf{ijo (e \dots)} && \textit{verb transitive}: to objectify \\ % no-dictionary
 && \\ % no-dictionary
%
\index{ilo}
\textbf{\dots ilo} && \textit{adjective}: useful \\ % no-dictionary
\textbf{\dots ilo} && \textit{adverb}: usefully \\ % no-dictionary
\textbf{ilo} && \textit{noun}: tool, device, machine, thing used for a specific purpose \\ % no-dictionary
 && \\ % no-dictionary
%
\index{jo}
\textbf{\dots jo} && \textit{adjective}: private, personal \\ % no-dictionary
\textbf{jo} && \textit{noun}: having, possessions, content \\ % no-dictionary
\textbf{jo (e \dots)} && \textit{verb transitive}: to have, to contain \\ % no-dictionary
 && \\ % no-dictionary
%
\index{kili}
\textbf{\dots kili} && \textit{adjective}: fruity \\ % no-dictionary
\textbf{\dots kili} && \textit{adverb}: fruity \\ % no-dictionary
\textbf{kili} && \textit{noun}: fruit, pulpy vegetable, mushroom \\ % no-dictionary
 && \\ % no-dictionary
%
\index{lukin}
\textbf{\dots lukin} && \textit{adjective}: visual(ly) \\ % no-dictionary
\textbf{\dots lukin} && \textit{adverb}: visual(ly) \\ % no-dictionary
\textbf{lukin} && \textit{noun}: view, look, glance, sight, gaze, glimpse, seeing, vision \\ % no-dictionary
\textbf{lukin} && \textit{verb intransitive}: to look, to watch out, to pay attention \\ % no-dictionary
\textbf{lukin (e \dots)} && \textit{verb transitive}: to see, to look at, to watch, to read \\ % no-dictionary
\textbf{lukin \dots} && \textit{auxiliary verb}: to seek to, try to, look for \\ % no-dictionary
 && \\ % no-dictionary
%
\index{ma}
\textbf{\dots ma} && \textit{adjective}: countrified, outdoor, alfresco, open-air \\ % no-dictionary
\textbf{ma} && \textit{noun}: land, earth, country, (outdoor) area \\ % no-dictionary
 && \\ % no-dictionary
%
\index{pakala}
\textbf{\dots pakala} && \textit{adjective}: destroyed, ruined, demolished, shattered, wrecked \\ % no-dictionary
\textbf{\dots pakala} && \textit{adverb}: destroyed, ruined, demolished, shattered, wrecked \\ % no-dictionary
\textbf{pakala} && \textit{noun}: blunder, accident, mistake, destruction, damage, breaking \\ % no-dictionary
\textbf{pakala} && \textit{verb intransitive}: to screw up, to fall apart, to break \\ % no-dictionary
\textbf{pakala (e \dots)} && \textit{verb transitive}: to screw up, to ruin, to break, to hurt, to injure, to damage \\ % no-dictionary
 && \\ % no-dictionary
%
\index{unpa}
\textbf{\dots unpa} && \textit{adjective}: erotic, sexual \\ % no-dictionary
\textbf{\dots unpa} && \textit{adverb}: erotic, sexual \\ % no-dictionary
\textbf{unpa} && \textit{noun}: sex, sexuality \\ % no-dictionary
\textbf{unpa} && \textit{verb intransitive}: to have sex \\ % no-dictionary
\textbf{unpa (e \dots)} && \textit{verb transitive}: to have sex with, to sleep with, to fuck \\ % no-dictionary
 && \\ % no-dictionary
%
\index{wile}
%\textbf{\dots wile} && \textit{adjective}: necessary \\ % no-dictionary
%\textbf{\dots wile} && \textit{adverb}: necessary \\ % no-dictionary
\textbf{wile} && \textit{noun}: desire, need, will \\ % no-dictionary
\textbf{wile (e \dots)} && \textit{verb transitive}: to want, need, wish, have to, must, will, should \\ % no-dictionary
\textbf{wile \dots} && \textit{auxiliary verb}: to want, need, wish, have to, must, will, should \\ % no-dictionary
\end{supertabular} \\
%
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Transitive Verbs, the Separator \textit{e} and Direct Objects}
%
\index{object!direct}
\index{verb!transitive}
\index{transitive verb}
\index{\textit{e}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We saw how phrases such as \textit{mi moku} could have two potential meanings. 
'I'm eating' or 'I am food'. 
There is one way to specify that you want to say. 

\begin{supertabular}{p{5,5cm}|ll}
mi moku e kili. && I eat fruit. \\
\end{supertabular} 

Also we discussed how \textit{sina pona}, like \textit{mi moku}, has two possible meanings. 
'You are good' or 'You're fixing'. 
Normally, it would mean 'You are good' simply because no one really says 'I'm fixing' without actually telling what it is that they are trying to fix. 

\begin{supertabular}{p{5,5cm}|ll}
ona li pona e ilo. && She's fixing the machine. \\
mi pona e ijo. && I'm fixing something. \\
\end{supertabular} 

Only a (composite) verb can stand in front of the \textit{e}. 
More specifically, it is a slot for a transitive verb. 
Transitive verbs are verbs after which a direct object (accusative object) can stand. 
A transitive verb does something to the direct object. 

The separator \textit{e} preface the direct object.
An object is an optional record supplement. 
A direct object is most strongly influenced by the action (i. e. the predicate).
Your can ask for direct object (accusative object) by' Who' or' What' (' What does she repair?'').
The direct object is part of the predicate phrase. 

In the direct object is the first slot after the separator \textit{e} always a noun or pronoun slot.
In the above examples the noun slots were filled with \textit{kili} and \textit{ijo}. 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Reflexive Pronouns}
%
\index{reflexive pronoun}
\index{pronoun!reflexive}
\index{pronoun!personal}
\index{\textit{mi}!reflexive pronoun}
\index{\textit{sina}!reflexive pronoun}
\index{\textit{ona}!reflexive pronoun}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A reflexive pronoun represents the subject in the direct object. 
So a slot for a reflective pronoun is located after the separator \textit{e}. 
In the following example, \textit{ona} is a reflexive pronoun, since it refers to the subject \textit{jan}. 

\begin{supertabular}{p{5,5cm}|ll}
jan li telo e ona. && A person washes himself. \\
\end{supertabular}

In this sentence the first \textit{mi} is a personal pronoun.
The \textit{mi} after the \textit{e} is a reflexive pronoun. 

\begin{supertabular}{p{5,5cm}|ll}
mi telo e mi. && I wash myself. \\
\end{supertabular}

Here a sentence with \textit{sina} as personal and reflective pronouns

\begin{supertabular}{p{5,5cm}|ll}
sina telo e sina. && You wash yourself. \\
\end{supertabular}

Here a sentence with \textit{ona} as personal and reflective pronouns

\begin{supertabular}{p{5,5cm}|ll}
ona telo e ona. && She washes herself. \\
\end{supertabular}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
\subsection*{Compound Sentences}
%
\index{sentence!compound}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
There are two ways to make compound sentences in Toki Pona; one way involves using \textit{li}, and the other way involves using \textit{e}. 
Since you've now studied both of these words, we'll cover how to use both of them to make compound sentences. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{'multiple_li'}
\subsubsection*{Several \textit{li} Separators for Several Predicate Phrases}
%
\index{predicate phrases!several}
\index{\textit{li}!several}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

It is possible to use the separator \textit{li} several times in a sentence. 
Each separator \textit{li} starts a new predicate phrase. 
This allows you to assign several actions or properties to one subject. 

\begin{supertabular}{p{5,5cm}|ll}
ona li ' pona li unpa. && He's awesome and has sex. \\
\end{supertabular} 

In the next example the separator \textit{li} is still omitted before \textit{moku} because the subject of the sentence is the personal pronoun \textit{mi}, we still use it before the second predicate, \textit{pakala}. 
Without the separator \textit{li} there, the sentence would be chaotic and confusing. 
Compound sentences with personal pronoun \textit{sina} as subject follow this same pattern. 

\begin{supertabular}{p{5,5cm}|ll}
mi moku li pakala. && I eat and destroy. \\
\end{supertabular} 

Predicate phrases are not nested. You can change the order. \\
ona li moku li ' pona. = ona li ' pona li moku. 

Each predicate phrase can of course contain direct objects. 

\begin{supertabular}{p{5,5cm}|ll}
mi moku e moku li lukin e ma. && I eat the food and look at the landscape. \\
\end{supertabular} 

The official Toki Pona book recommends to use only one predicate phrase for the personal pronouns \textit{mi} or \textit{sina} as subject. 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\label{'multiple_e'}
\subsubsection*{Several \textit{e} Separators for Several direct Objects}
%
\index{object!several}
\index{\textit{e}!several}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
For the other type of compound sentences, one predicate phrase has several direct objects. 
In other words, the action of a transitive verb refers to several things. 

\begin{supertabular}{p{5,5cm}|ll}
mi moku e kili e telo. && I eat/drink fruit and water. \\
mi wile lukin e ma e suno. && I want to see the land and the sun. \\
\end{supertabular}
 
\textit{e} phrases are not nested. You can change the order. \\
mi moku e moku e telo. = mi moku e telo e moku. 

We can combine several \textit{li} and \textit{e}.
We have two predicate phrases with two direct objects each. 
However, it is better to use several short sentences. 

\begin{supertabular}{p{5,5cm}|ll}
mi moku e kili e telo li lukin e ma e jan. I eat fruits and water and see land and people. \\
\end{supertabular} 

%
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Practice (Answers: Page~\pageref{'direct_objects_compund_sentences'})}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Please write down your answers and check them afterwards. 

\begin{supertabular}{p{5,5cm}|ll}
How to ask for the direct object? &&  \\ % no-dictionary
What word type has a predicate before the separator \textit{e}?  &&  \\ % no-dictionary
To which phrase in the sentence belongs a direct object? &&  \\ % no-dictionary
What kinds of words are possible after the separator \textit{e}?  &&  \\ % no-dictionary
What is a predicate noun?  &&  \\ % no-dictionary
Where are possible slots for reflexive pronouns? &&  \\ % no-dictionary
Is it possible to describe several properties of a subject with several \textit{e}? &&  \\ % no-dictionary
How can you create multiple predicate phrases in a sentence?  &&  \\ % no-dictionary
\end{supertabular} 

Try to translate these sentences. 
You can use the tool \textit{Toki Pona Parser} (\cite{www:rowa:02}) for spelling and grammar check. 

\begin{supertabular}{p{5,5cm}|ll}
I have a tool. &&  \\ % no-dictionary
She's eating fruit. &&  \\ % no-dictionary
Something is watching me. &&  \\ % no-dictionary
Pineapple is a food and is good. &&  \\ % no-dictionary
He washes himself. &&  \\ % & English - Toki Pona
  && \\ % no-dictionary
mi ' jan li ' suli. &&  \\ % no-dictionary
\end{supertabular} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% eof
