%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Gender,Unofficial Words,Addressing People,\\Interjections, Commands} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vocabulary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\index{\textit{a}}
\index{\textit{awen}}
\index{\textit{mama}}
\index{\textit{mije}}
\index{\textit{meli}}
\index{\textit{mu}}
\index{\textit{nimi}}
\index{\textit{o}}
%\index{\textit{pata}}
\index{\textit{pona}}
\index{\textit{toki}}
\index{wait}
\index{pause}
\index{remaining}
\index{parent}
\index{man}
\index{husband}
\index{boyfriend}
\index{male}
\index{woman}
\index{wife}
\index{girlfriend}
\index{female}
\index{sound!animal}
\index{animal!sound}
\index{name}
\index{word}
\index{vocative}
\index{imperative}
%\index{sibling}
\index{yay}
\index{cool}
\index{good}
\index{hey}
\index{language}
\begin{supertabular}{p{2,5cm}|ll}
\textbf{a} && \textit{interjection}: ah, ha, uh, oh, ooh, aw, well (emotion word) \\ % no-dictionary
\textbf{a a a!} && \textit{interjection}: laugh \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots awen} && \textit{adjective}: remaining, stationary, permanent, sedentary \\ % no-dictionary
\textbf{\dots awen} && \textit{adverb}: still, yet \\ % no-dictionary
\textbf{awen} && \textit{noun}: inertia, continuity, continuum, stay \\ % no-dictionary
\textbf{awen} && \textit{verb intransitive}: to stay, to wait,to remain \\ % no-dictionary
\textbf{awen (e \dots)} && \textit{verb transitive}: to keep \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots mama} && \textit{adjective}: of the parent, parental, maternal, fatherly, motherly, mumsy \\ % no-dictionary
\textbf{mama} && \textit{noun}: parent, mother, father \\ % no-dictionary
\textbf{mama (e \dots)} && \textit{verb transitive}: to mother sb., to wet-nurse, mothering \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots meli} && \textit{adjective}: female, feminine, womanly \\ % no-dictionary
\textbf{meli} && \textit{noun}: woman, female, girl, wife, girlfriend \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots mije} && \textit{adjective}: male, masculine, manly \\ % no-dictionary
\textbf{mije} && \textit{noun}: man, male, husband, boyfriend \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots mu} && \textit{adjective}: animal nois- \\ % no-dictionary
\textbf{\dots mu} && \textit{adverb}: animal nois- \\ % no-dictionary
\textbf{mu!} && \textit{interjection}: woof! meow! moo! etc. (cute animal noise) \\ % no-dictionary
\textbf{mu} && \textit{noun}: animal noise \\ % no-dictionary
\textbf{mu (e \dots)} && \textit{verb transitive}: to make animal nois \\ % no-dictionary
 && \\ % no-dictionary
\textbf{nimi} && \textit{noun}: word, name \\ % no-dictionary
\textbf{nimi (e \dots )} && \textit{verb transitive}: to name \\ % no-dictionary
 && \\ % no-dictionary
\textbf{o!} && \textit{interjection}: hey! (calling somebody's attention) \\ % no-dictionary
\textbf{\dots o, \dots} && \textit{interjection}: adressing people \\ % no-dictionary
\textbf{o \dots !} && \textit{subject}: An 'o' is used for imperative (commands). 'o' replace the subject.  \\ % no-dictionary
\textbf{\dots o \dots !} && \textit{separator}: An 'o' is used for imperative (commands): 'o' replace 'li'. \\ % no-dictionary
 && \\ % no-dictionary
\textbf{ala!} && \textit{interjection}: no! \\ % no-dictionary
 && \\ % no-dictionary
\textbf{ike!} && \textit{interjection}: oh dear! woe! alas! \\ % no-dictionary
 && \\ % no-dictionary
\textbf{jaki!} && \textit{interjection}: ew! yuck! \\ % no-dictionary
 && \\ % no-dictionary
\textbf{pakala!} && \textit{interjection}: damn! fuck! \\ % no-dictionary
 && \\ % no-dictionary
\textbf{pona!} && \textit{interjection}: great! good! thanks! OK! cool! yay! \\ % no-dictionary
 && \\ % no-dictionary
\textbf{toki!} && \textit{interjection}: hello, hi, good morning, \\ % no-dictionary
 && \\ % no-dictionary
\textbf{"} && \textit{separator}: Quotation marks are used for words with original spelling or for quotes. \\ % no-dictionary
\end{supertabular} \\
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{gender}
\subsection*{Gender}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Toki Pona doesn't have any grammatical gender like in most Western languages.  
However, some words in Toki Pona (such as \textit{mama}) don't tell you which gender a person is, and so we use \textit{mije} and \textit{meli} to distinguish. 

\index{mother}
\index{father}
\index{sister}
\index{brother}
\begin{supertabular}{p{5,5cm}|ll}
mama && a parent in general (mother or father) \\
mama meli && mother \\
mama mije && father \\
% pata && a sibling (brother or sister) \\
% pata meli && sister \\
% pata mije && brother \\
\end{supertabular} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{unofficial words}
\index{words!unofficial}
\index{nation!name of}
\index{country!name of}
\index{languages!name of}
\index{religion!name of}
\index{person!name of}
\index{woman!name of}
\index{man!name of}
\index{name!of a nation}
\index{name!of a country}
\index{name!of a languages}
\index{name!of a religion}
\index{name!of a person}
\index{name!of a woman}
\index{name!of a man}
\subsection*{Unofficial words}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Take a moment to look over the Toki Pona dictionary in the Appendix (Page~\pageref{'unofficial_words'}). 
There are no words for the names of nations; there are also aren't any words for religions or even other languages. 
The reason that these words aren't in the dictionary is because they are 'unofficial'. 
Before using an unofficial word, we often adapt the word to fit into Toki Pona's phonetic rules. 
So, for example, America becomes \textit{Mewika}, Canada becomes \textit{Kanata}, and so on. 
Now the thing about these unofficial words is that they can never be used by themselves. 
They are always treated like adjectives, and so they have to be used with a noun. 

\begin{supertabular}{p{5,5cm}|ll}
ma Kanata li ' pona. && Canada is good. \\
ma Italija li ' pona lukin. && Italy is beautiful. \\
mi wile tawa ma Tosi. && I want to go to Germany. \\
\end{supertabular} 

\index{city}
Do you remember that \textit{ma tomo} means 'city'? 

\begin{supertabular}{p{5,5cm}|ll}
ma tomo Lantan li ' suli. && London is big. \\
ma tomo Pelin && Berlin \\
ma tomo Alenta && Atlanta \\
ma tomo Loma && Rome \\
mi kama tan ma tomo Pelin. && I'm from Berlin. \\
\end{supertabular} 

If you want to talk about a language, you simply use \textit{toki} and then attach the unofficial word onto it. 

\begin{supertabular}{p{5,5cm}|ll}
toki Inli li ' pona. && The English language is good. \\
ma Inli li ' pona. && England is good. \\
toki Kanse && French language \\
toki Epelanto && Esperanto \\
\end{supertabular} 

If you want to talk about a person who is from a certain place, you just say \textit{jan} and then attach the unofficial word at the end, like this. 

\begin{supertabular}{p{5,5cm}|ll}
jan Kanata && Canadian person \\
jan Mesiko && Mexican person \\
\end{supertabular} 

Of course, you can use other nouns to describe certain people, too. 
For example, you can use the \textit{mije} and \textit{meli} that you learned a few minutes ago. 

\begin{supertabular}{p{5,5cm}|ll}
meli Italija && Italian woman \\
mije Epanja && Spanish man \\
\end{supertabular} 

Now suppose you want to talk about someone using their name. 
For example, what if you want to say 'Lisa is cool'? 
To say a person's name in Toki Pona, you just say \textit{jan} and then the person's name. 

\begin{supertabular}{p{5,5cm}|ll}
jan Lisa li ' pona. && Lisa is cool. \\
\end{supertabular} 

Like for the names of countries, we often adapt a person's name to fit into Toki Pona's phonetic rules. 
Keep in mind that no one is going to pressure you to adopt a tokiponized name; it's just for fun. 

\begin{supertabular}{p{5,5cm}|ll}
jan Pentan li pana e sona, tawa mi. && Brandon teaches to me. \\
jan Mewi li toki, tawa mi. && Mary's talking to me. \\
jan Nesan li ' musi. && Nathan is funny. \\
jan Eta li ' jan unpa. && Heather is a whore. \\
pana e sona && to teach (give knowledge) \\
\end{supertabular} 

This is the way to say your name. 

\begin{supertabular}{p{5,5cm}|ll}
mi ' jan Pepe. && I am Pepe. \\
\end{supertabular} 

Nobody is forcing you to use a name in Toki Pona style.
This is pure fun.
If you use the original spelling, please put it in quotation marks.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
%\subsection*{Addressing People, Commands, Interjections}
\index{addressing!\textit{o}}
\index{\textit{o}!addressing people}
\index{addressing people}
\index{person!addressing}
\index{person!attention}
\index{attention!a person}
\subsection*{Addressing People}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Sometimes you need to get a person's attention before you can talk to him. 
When you want to address someone like that before saying the sentence, you just follow this same pattern. 
\textit{jan} (name) \textit{o} (sentence). 

\begin{supertabular}{p{5,5cm}|ll}
jan Ken o, pipi li lon len sina. && Ken, a bug is on your shirt. \\
jan Keli o, sina pona lukin. && Kelly, you are pretty. \\
jan Mawen o, sina wile ala wile moku? && Marvin, are you hungry? \\
jan Tepani o, sina ike tawa mi. && Steffany, I don't like you. \\
\end{supertabular} 

Although it's not essential, you should try to remember to put the comma after the \textit{o}. 
% A comma in Toki Pona is only used after the word \textit{o} and only to addressing people. 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{command}
\index{command!\textit{o}}
\index{\textit{o}!command}
\subsection*{Commands}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
You simply say \textit{o} and then whatever you want the person to do. 
An imperative sentence (command) ends with an exclamation mark.

\begin{supertabular}{p{5,5cm}|ll}
o pali! && Work! \\
o awen! && Wait! \\
o lukin e ni! && Watch this! \\
o tawa ma tomo, poka jan pona sina! && Go to the city with your friend! \\
\end{supertabular} 

We've learned how to address people and how to make commands; now let's put these two concepts together. 
Suppose you want to address someone and tell them to do something. 
Notice how one of the \textit{o}'s got dropped, as did the comma. 

\begin{supertabular}{p{5,5cm}|ll}
jan San o, ...  && John, ... \\
 ... o tawa tomo sina! && ... go to your house! \\
jan San o tawa tomo sina!  && John, go to your house! \\
 && \\ % no-dictionary
jan Ta o toki ala, tawa mi! && Todd, don't talk to me! \\
jan Sesi o moku e kili ni! && Jessie, eat this fruit!. \\
\end{supertabular} 

The \textit{o} replaces the \textit{li}. After \textit{mi}  and \textit{sina} it is also used.

\begin{supertabular}{p{5,5cm}|ll}
sina o telo e sina! && Wash yourself! \\
\end{supertabular} 

\index{let's go}
This structure can also be used to make sentences like 'Let's go'.

\begin{supertabular}{p{5,5cm}|ll}
mi mute o tawa! && Let's go. \\
mi mute o ' musi! && Let's have fun. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
\index{interjection}
\index{interjection!of an animal}
\index{interjection!laughter}
\index{laughter}
\index{Hahaha}
\index{Hello}
\index{f-ck}
\index{d-mm}
\index{cool}
\subsection*{Interjections}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
An interjection sentence makes also conveys excitement and ends with an exclamation mark.
I'm going to split the interjections into two groups so that they're easier to explain. 
The first group is very straight-forward.

\begin{supertabular}{p{5,5cm}|ll}
jan Lisa o, toki! && Hello Lisa! \\
pona! && Yay! Good! Hoorah! \\
ike! && Oh no! Uh! oh! Alas! \\
pakala! && F-ck! D-mn! \\
mu! && Meow! Woof! Grrr! Moo! (sounds made by animals) \\
a! && Ooh, Ahh! Unh! Oh! \\
a a a! && Hahaha! (laughter) \\
\end{supertabular} 

\textit{a} adds emotion or emphasis. It can be used at the end of a sentence.

\begin{supertabular}{p{5,5cm}|ll}
sina suli a! && You are so tall! \\
\end{supertabular} 

\textit{o} and \textit{a} only used when the person makes you feel really emotional. 
For example, you might use the \textit{a} when you haven't seen the person for a very long time or when having sex.  

\begin{supertabular}{p{5,5cm}|ll}
jan Epi o a! && Oh Abbie! \\
\end{supertabular} 

The second group of interjections are kind like salutations.

\index{Good day}
\index{Sleep well}
\index{Enjoy your meal}
\index{Bye}
\index{Good bye}
\index{Welcome}
\index{Have fun}
\begin{supertabular}{p{5,5cm}|ll}
toki! && Hello!, Hi! \\
suno pona! && Good sun! Good day! \\
lape pona! && Sleep well! Have a good night! \\
moku pona! && Good food! Enjoy your meal! \\
mi tawa && I'm going. Bye! \\
tawa pona! && (in reply) Go well! Good bye! \\
kama pona! && Come well! Welcome! \\
musi pona! && Good fun! Have fun! \\
\end{supertabular}  
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Practice 9 (Answers: Page~\pageref{'gender_unofficial_words_etc'})}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Try to translate these sentences. 
You can use the tool \textit{Toki Pona Parser} (\cite{www:rowa:02}) for spelling and grammar check. 

\begin{supertabular}{p{5,5cm}|ll}
Susan is crazy. &&   \\ % no-dictionary
Go!  &&  \\ % no-dictionary
Mama, wait!  &&  \\ % no-dictionary
I come from Europe. * &&  \\ % no-dictionary
Hahaha! That's funny!  &&  \\ % no-dictionary
My name is Ken. &&  \\  % no-dictionary
Hello Lisa!  && \\  % no-dictionary
F-ck!  &&   \\ % no-dictionary
I want to go to Australia. **  &&  \\  % no-dictionary
Bye! *** &&   \\ % no-dictionary
 && \\ % no-dictionary
mu! &&   \\ % no-dictionary
mi wile kama sona e toki Inli. **** &&  \\ % no-dictionary
jan Ana o pana e moku tawa mi!  &&  \\ % no-dictionary
o tawa musi, poka mi! ***** &&  \\ % no-dictionary
jan Mose o lawa e mi mute, tawa ma pona! &&   \\ % no-dictionary
tawa pona!  &&   \\ % no-dictionary
\end{supertabular} 

* Europe = Elopa \\
** Australia = Oselija \\
*** spoken by the person who's leaving \\
****  \textit{kama sona} = learn. This is similar to how \textit{kama jo} = 'get', like we learned in lesson six. \\
% ***** If you've forgotten what \textit{tawa musi} means, we learned it in lesson eight. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% eof
