%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Sonja Lang created the language Toki Pona in the year 2001. 
Her aim was minimalism. 
Toki Pona consists of only about 120 words, which are never altered. 
The part of speech signified by a word varies in accordance with its position in the sentence. 
To describe things in more detail you have to combine words.

It is not the goal of Toki Pona to describe complex issues. 
Dissertations and scientific papers will never be written in Toki Pona. 
Lawyers, bureaucrats, theologians and politicians should be warned of the side-effects of this language.

It is not the aim of Toki Pona to solve the communication problems of the world. 
But you can learn this language in a month. 
Toki Pona is easy in an intelligent way and it's yoga for the brain. 
People who hate nested subordinate clauses and commas will certainly have fun with Toki Pona.

Maybe only one natural language can be compared to Toki Pona. 
It is the language of the Pirah\'{a} (\cite{www:piraha:01}). 
For example, this language has no recursion. 

%
%% 
%
Toki Pona has evolved since 2001. 
Therefore these lessons are based on the tutorials from B.\ J.\ Knight (jan Pije) \cite{www:Pije:01} (2003) and the official Toki Pona book \cite{www:tokipona.org} by Sonja Lang (2014). 
But I tried not to carry over mistakes and inaccuracies. 
In my lessons, great importance is attached to the presentation of grammatical rules. 
This avoids misunderstandings due to incorrect grammar.

So have fun with the lessons and with learning Toki Pona. 
Memrise helps for learning vocabulary \cite{www:memrise:01}. Links related to Toki Pona can be found on the website \cite{www:rowa:01}. 
A dictionary can be found here \cite{www:rowa:01}. 

You can use the tool \textit{Toki Pona Parser} (\cite{www:rowa:02}) for checking the spelling, grammar and ambiguity of Toki Pona sentences. 


\textit{toki pona li ' pona, tawa sina.}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% eof
