%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Living Things}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vocabulary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{supertabular}{p{2,5cm}|ll}
%
\index{akesi}
\textbf{\dots akesi} && \textit{adjective}: amphibian-, reptilian-, slimy \\ % no-dictionary
\textbf{akesi} && \textit{noun}: reptile, amphibian; non-cute animal \\ % no-dictionary
 && \\ % no-dictionary
%
\index{alasa}
\textbf{\dots alasa} && \textit{adjective}: hunting-, -hunting, hunting \\ % no-dictionary
\textbf{alasa} && \textit{noun}: hunting \\ % no-dictionary
\textbf{alasa (e \dots)} && \textit{verb transitive}: to hunt, to forage \\ % no-dictionary
 && \\ % no-dictionary
%
\index{kala}
\textbf{\dots kala} && \textit{adjective}: fish- \\ % no-dictionary
\textbf{kala} && \textit{noun}: fish, marine animal, sea creature \\ % no-dictionary
 && \\ % no-dictionary
%
\index{kasi}
\textbf{\dots kasi} && \textit{adjective}: vegetable, vegetal, biological, biologic, leafy \\ % no-dictionary
\textbf{kasi} && \textit{noun}: plant, vegetation, herb, leaf \\ % no-dictionary
\textbf{kasi} && \textit{verb intransitive}: to grow \\ % no-dictionary
\textbf{kasi (e \dots)} && \textit{verb transitive}: to plant, to grow \\ % no-dictionary
 && \\ % no-dictionary
%
\index{moli}
\textbf{\dots moli} && \textit{adjective}: dead, dying, fatal, deadly, lethal, mortal, deathly, killing \\ % no-dictionary
\textbf{\dots moli} && \textit{adverb}: mortally \\ % no-dictionary
\textbf{moli} && \textit{noun}: death, decease \\ % no-dictionary
\textbf{moli} && \textit{verb intransitive}: to die, to be dead \\ % no-dictionary
\textbf{moli (e \dots)} && \textit{verb transitive}: to kill \\ % no-dictionary
kama \textbf{moli} && \textit{intransitives Verb}: dieing \\ % no-dictionary
 && \\ % no-dictionary
%
\index{monsuta}
\textbf{\dots monsuta} && \textit{adjective}: fearful, afraid \\ % no-dictionary
\textbf{monsuta} && \textit{noun}: monster, monstrosity, fearful thing, fright, mythical creatures, fear \\ % no-dictionary
 && \\ % no-dictionary
%
\index{namako}
\textbf{\dots namako} && \textit{adjective}: spicy, piquant \\ 
\textbf{namako} && \textit{noun}: spice, something extra, food additive, accessory \\ % no-dictionary
\textbf{namako (e \dots)} && \textit{verb transitive}: to spice, to flavor, to decorate \\ % no-dictionary
 && \\ % no-dictionary
%
\index{pan}
\textbf{pan} && \textit{noun}: cereal, grain; barley, corn, oat, rice, wheat; bread, pasta \\ % no-dictionary
\textbf{pan (e \dots)} && \textit{verb transitive}: to sow \\ % no-dictionary
 && \\ % no-dictionary
%
\index{soweli}
\textbf{\dots soweli} && \textit{adjective}: animal \\ % no-dictionary
\textbf{soweli} && \textit{noun}: animal, especially land mammal, lovable animal, beast \\ % no-dictionary
 && \\ % no-dictionary
%
\index{waso}
\textbf{\dots waso} && \textit{adjective}: bird- \\ % no-dictionary
\textbf{waso} && \textit{noun}: bird, bat; flying creature, winged animal \\ % no-dictionary
%
\end{supertabular} \\
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
%
\subsection*{Ambiguity of Names of Living Things}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{\textit{soweli}}
%
\index{\textit{soweli}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{soweli} is basically for all types of mammals. 
If you're eating meat that comes from a \textit{soweli} (such as beef or pork), it is still called \textit{soweli}. 

\begin{supertabular}{p{5,5cm}|ll}
soweli ni li ' pona moku. && This cow is good to eat. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{\textit{waso}}
%
\index{\textit{waso}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{waso} includes all birds and flying animals. 

\begin{supertabular}{p{5,5cm}|ll}
mi wile moku e waso. && I want to eat chicken. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsubsection*{\textit{akesi}}
%
\index{\textit{akesi}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{akesi} covers all of the reptiles, amphibians, dinosaurs and monsters.

\begin{supertabular}{p{5,5cm}|ll}
akesi pi telo moli && venomous snakes, poisonous frogs \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsubsection*{\textit{kala}}
%
\index{\textit{kala}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{kala} is used to talk about fish and other animals that live in water. 
Also, \textit{kala} can be used as an adjective. 

\begin{supertabular}{p{5,5cm}|ll}
jan kala && mermaid. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsubsection*{\textit{pipi}}
%
\index{\textit{pipi}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{pipi} is used for all types of bugs (spiders, ants, roaches, butterflies, ...). 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
%
\subsubsection*{\textit{kasi}}
%
\index{\textit{kasi}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{kasi} is used to talk about all plants and plant-like things. 

\begin{supertabular}{p{5,5cm}|ll}
kasi kule && flower \\
kasi suli && trees, big shrubs \\
kasi anpa && grass \\
ma kasi && forest, jungle \\
kasi nasa / kasi sona && hemp  \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\subsubsection*{\textit{pan} and \textit{namako}}
%
\index{\textit{pan}}
\index{\textit{namako}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supertabular}{p{5,5cm}|ll}
pan ni li ' moku ike. && This pasta is unappetizing. \\
o pana e namako, tawa mi. &&   Give me some spice. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsubsection*{\textit{moli}}
%
\index{\textit{moli}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
It can be used to say that something is dead, dying or killing.

\begin{supertabular}{p{5,5cm}|ll}
pipi li ' moli. && The bug is dead. \\
soweli li kama moli. && The cheetah is dying. \\
jan li moli e waso. && The person killed the bird. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsubsection*{\textit{alasa}}
%
\index{\textit{alasa}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{supertabular}{p{5,5cm}|ll}
jan li alasa e soweli. && Somebody hunt a buffalo. \\
jan alasa pona li ' wawa. && A good hunter is strong. \\
\end{supertabular}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsubsection*{\textit{monsuta}}
%
\index{\textit{monsuta}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{supertabular}{p{5,5cm}|ll}
monsuta waso pi pan linja li pali e ali. &&  The Flying Spaghetti Monster has created the world. \\
\end{supertabular}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
%
\subsection*{Practice (Answers: Page~\pageref{'living_things'})}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Please write down your answers and check them afterwards. 

Try to translate these sentences. 
You can use the tool \textit{Toki Pona Parser} (\cite{www:rowa:02}) for spelling and grammar check. 

\begin{supertabular}{p{5,5cm}|ll}
I want a puppy. &&   \\ % no-dictionary
Ahh! The dinosaur wants to eat me!  &&  \\ % no-dictionary
The mosquito bit me.  &&  \\ % no-dictionary
Cows say moo.  &&  \\ % no-dictionary
Birds fly in air. * &&  \\ % no-dictionary
Let's eat fish.  &&  \\ % no-dictionary
Flowers are pretty. ** &&  \\ % no-dictionary
I like plants. &&   \\ % no-dictionary
 && \\ % no-dictionary
mama ona li kepeken kasi nasa.  &&  \\ % no-dictionary
akesi li pana e telo moli. &&  \\  % no-dictionary
pipi li moku e kasi.  &&  \\ % no-dictionary
soweli mi li kama moli.  &&  \\ % no-dictionary
jan Pawe o, mi wile ala moli. && \\   % no-dictionary
mi lon ma kasi. &&   \\ % no-dictionary
\end{supertabular} 

* Think: 'Birds go in air.' \\
** Think: 'Colorful plants are good to see.' 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% eof
