%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negation, Yes/No Questions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vocabulary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{supertabular}{p{2,5cm}|ll}
%
\index{ala}
\textbf{\dots ala} && \textit{adjective}: no, not, none, un- \\ % no-dictionary
\textbf{\dots ala} && \textit{adverb}: don't \\ % no-dictionary
\textbf{ala} && \textit{noun}: nothing, negation, zero \\ % no-dictionary
 && \\ % no-dictionary
%
\index{ken}
\textbf{ken} && \textit{noun}: possibility, ability, power to do things, permission \\ % no-dictionary
\textbf{ken} && \textit{verb intransitive}: can, is able to, is allowed to, may, is possible \\ % no-dictionary
\textbf{ken \dots} && \textit{auxiliary verb}: to can, may \\ % no-dictionary
\textbf{ken (e \dots)} && \textit{verb transitive}: to make possible, to enable, to allow, to permit \\ % no-dictionary
 && \\ % no-dictionary
%
\index{lape}
\textbf{\dots lape} && \textit{adjective}: sleeping, of sleep, dormant \\ % no-dictionary
\textbf{\dots lape} && \textit{adverb}: asleep \\ % no-dictionary
\textbf{lape} && \textit{noun}: sleep, rest \\ % no-dictionary
\textbf{lape} && \textit{verb intransitive}: to sleep, to rest \\ % no-dictionary
\textbf{lape (e \dots)} && \textit{verb transitive}: to knock out \\ % no-dictionary
 && \\ % no-dictionary
%
\index{musi}
\textbf{\dots musi} && \textit{adjective}: artful, fun, recreational \\ % no-dictionary
\textbf{\dots musi} && \textit{adverb}: cheerfully \\ % no-dictionary
\textbf{musi} && \textit{noun}: fun, playing, game, recreation, art, entertainment \\ % no-dictionary
\textbf{musi} && \textit{verb intransitive}: to play, to have fun \\ % no-dictionary
\textbf{musi (e \dots)} && \textit{verb transitive}: to amuse, to entertain \\ % no-dictionary
 && \\ % no-dictionary
%
\index{pali}
\textbf{\dots pali} && \textit{adjective}: active, work-related, operating, working \\ % no-dictionary
\textbf{\dots pali} && \textit{adverb}: actively, briskly \\ % no-dictionary
\textbf{pali} && \textit{noun}: activity, work, deed, project \\ % no-dictionary
\textbf{pali} && \textit{verb intransitive}: to act, to work, to function \\ % no-dictionary
\textbf{pali (e \dots)} && \textit{verb transitive}: to do, to make, to build, to create \\ % no-dictionary
 && \\ % no-dictionary
%
\index{wawa} 
\textbf{\dots wawa} && \textit{adjective}: energetic, strong, fierce, intense, sure, confident \\ % no-dictionary
\textbf{\dots wawa} && \textit{adverb}: strongly, powerfully \\ % no-dictionary
\textbf{wawa} && \textit{noun}: energy, strength, power \\ % no-dictionary
\textbf{wawa (e \dots)} && \textit{verb transitive}: to strengthen, to energize, to empower \\ % no-dictionary
\end{supertabular} \\
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
%
\subsection*{Negation}
%
\index{negation}
\index{\textit{ala}!negation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsubsection*{The Adverb \textit{ala}}
%
\index{\textit{ala}!adverb}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In English, you make a verb negative by adding 'not' in front of the verb.
In Toki Pona you put the adverb \textit{ala} after the verb. 

\index{lazy}
\begin{supertabular}{p{5,5cm}|ll}
mi lape ala. && I'm not sleeping. \\ 
mi musi ala. && I'm not having fun. / I'm bored. \\
mi wawa ala. && I'm not strong. / I'm weak. \\
mi wile ala tawa musi. && I don't want to dance. \\
tawa musi && dance (move entertainingly) \\
mi wile ala pali. && I'm lazy. \\
\end{supertabular} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The Adjective \textit{ala}}
%
\index{\textit{ala}!adjective}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{supertabular}{p{5,5cm}|ll}
jan ala li toki. && Nobody is talking. \\
\end{supertabular} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The Noun \textit{ala}}
%
\index{\textit{ala}!noun}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

\begin{supertabular}{p{5,5cm}|ll}
ala li ' jaki. && Nothing is dirty. \\
\end{supertabular} 

%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
\subsection*{Yes/No Questions with \textit{ala}}
%
\index{question!yes,no}
\index{yes,no!question}
%

Yes/No questions are formed in \textit{toki pona} according to a simple pattern.
\textit{ala} is appended to the part of the sentence that is asked for and the part of the sentence is repeated.
As a rule, this part of the sentence is the entire predicate (sentence statement).
The rest of the sentence structure does not change.
A question is ended with a question mark.

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{An Intransitive Verb as Subject of the Question}
%

If the yes/no question refers to an intransitive verb, the adverb \textit{ala} is appended to it and the intransitive verb is repeated.
Consider the following example.

\begin{supertabular}{p{5,5cm}|ll}
sina tawa mi. &&  You're leaving me. \\
\end{supertabular} 

If we want to ask 'Are you leaving me?', we append the adverb \textit{ala} to the intransitive verb \textit{tawa}.
Then we repeat the intransitive verb \textit{tawa}.

\begin{supertabular}{p{5,5cm}|ll}
sina tawa ala tawa mi? &&  Are you leaving me? \\
\end{supertabular}

Here are more examples.

\begin{supertabular}{p{5,5cm}|ll}
ona li lon ala lon tomo? && Is he in the house? \\
sina kepeken ala kepeken ilo? && Are you using the tool? \\
pona li kama ala kama?  && Is the good coming? \\
sina sewi ala sewi? && Are you getting up? \\
\end{supertabular} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{A Transitive Verb as Subject of the Question}
%

If the yes/no question refers to a transitive verb, the adverb \textit{ala} is appended to it and the transitive verb is repeated.

\begin{supertabular}{p{5,5cm}|ll}
sina pona ala pona e ilo? && Are you fixing the tool? \\
sina pana ala pana e moku tawa jan lili? && Did you give food to the child? \\
pipi li moku ala moku e kili? && Are the bugs eating the fruit? \\
ona li mama ala mama e sina? && Does she mother you? \\
\end{supertabular} 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{An auxiliary verb as Subject of the Question}
%
As we have learned, the auxiliary verb and the verb together form the predicate. 
If the yes/no question refers to an auxiliary verb, then the adverb \textit{ala} is not attached to the predicate, but directly to the auxiliary verb.
Only the auxiliary verb is repeated. 
Then the verb follows. 

\begin{supertabular}{p{5,5cm}|ll}
sina wile ala wile moku? && Do you want to eat? \\ 
sina ken ala ken lape? && Can you sleep? \\
sina kama ala kama jo e pali ni? && Did you get this job?
\end{supertabular} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{An Predicate Noun as Subject of the Question}
%
We had already learned the difference between verb and predicate. 
In Toki Pona are sentences without verb possible. 
Then nouns serve as predicate nouns or adjectives as predicate adjectives. 

In the lessons of B. J. Knight (2003) and the official Toki Pona book of Sonja Lang \cite{www:tokipona.org} yes/no questions with \textit{ala} are defined only with verbs. 
But this contradicts their own examples as well as common practice. 
For example one cannot formulate the question 'Is she a mother? 
In these lessons I will therefore not adhere to this limitation. 

If the yes/no question refers to a predicate noun, the adjective \textit{ala} is added to it and the predicate noun is repeated.

\begin{supertabular}{p{5,5cm}|ll}
ona li ' mama ala ' mama ? && Is she a parent? \\
ni li ' jan ala ' jan? && Is this a person? \\
ni li ' kili ala ' kili? && Is this a banana? \\
ni li ' tomo pi telo nasa ala ' tomo pi telo nasa? && Is this a pub? \\
ona li ' jan pi pona lukin ala ' jan pi pona lukin? && Is she an attractive person? \\
\end{supertabular} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{An Predicate Adjective as Subject of the Question}
%

If the yes/no question refers to a predicate adjective, the adjective \textit{ala} is added to it and the predicate adjective is repeated.

\begin{supertabular}{p{5,5cm}|ll}
sina ' pona ala ' pona? && Are you OK? \\
mi ' pona ala ' pona, tawa sina? && Do you like me? \\
suno li ' suli ala ' suli? && Is the sun big? \\
len sina li ' telo ala ' telo? && Are your clothes wet? \\
\end{supertabular} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\subsection*{Yes/No Answering}
%
\index{answering!yes,no}
\index{yes}
\index{no}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

If you want to say 'yes', you simply repeat the predicate or the auxiliary verb of the sentence. 
If you want to say 'no', you repeat the predicate or the auxiliary verb and add \textit{ala}} after it. 

\begin{supertabular}{p{5,5cm}|ll}
sina wile ala wile moku? && Do you want to eat? \\ 
wile && Yes. \\ % no-dictionary
wile ala && No. \\ % no-dictionary
 && \\ % no-dictionary
sina lukin ala lukin e kiwen? && Do you see the rock? \\ 
lukin && Yes. \\ % no-dictionary
lukin ala && No. \\ % no-dictionary
 && \\ % no-dictionary
sina sona ala sona e toki mi? && Do you understand what I'm saying? \\ 
sona && Yes. \\ % no-dictionary
sona ala && No. \\ % no-dictionary
 && \\ % no-dictionary
ni li ' tomo pi telo nasa ala ' tomo pi telo nasa? && Is this a pub? \\
tomo pi telo nasa.  && Yes. \\  % no-dictionary
tomo pi telo nasa ala. && No. \\ % no-dictionary
\end{supertabular}  

%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
%
\subsection*{Practice (Answers: Page~\pageref{'negation_yes_no_questions'})}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Please write down your answers and check them afterwards. 

\begin{supertabular}{p{5,5cm}|ll}
Which separator is at the end of a question? &&  \\ % no-dictionary
How is a verb negated in Toki Pona? &&  \\ % no-dictionary
How do you answer in Toki Pona negative to a yes/no question? &&  \\ % no-dictionary
How do you answer positively to a yes/no question in Toki Pona? &&  \\ % no-dictionary
\end{supertabular} 

Try to translate these sentences. 
You can use the tool \textit{Toki Pona Parser} (\cite{www:rowa:02}) for spelling and grammar check. 

\begin{supertabular}{p{5,5cm}|ll}
You have to tell me why. *  &&   \\ % no-dictionary
Is a bug beside me?  &&    \\ % no-dictionary
I can't sleep.  &&    \\ % no-dictionary
I don't want to talk to you.  &&    \\ % no-dictionary
He didn't go to the lake.   &&   \\ % no-dictionary
\end{supertabular}

\begin{supertabular}{p{5,5cm}|ll}
sina wile ala wile pali? wile ala.  &&    \\ % no-dictionary
jan utala li seli ala seli e tomo?   &&   \\ % no-dictionary
jan lili li ken ala moku e telo nasa.   &&   \\ % no-dictionary
sina kepeken ala kepeken ni?  &&    \\ % no-dictionary
sina ken ala ken kama?   &&   \\ % no-dictionary
sina pona ala pona? &&   \\ % no-dictionary
\end{supertabular} 

* Think: 'You have to tell the reason to me.' 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% eof
