%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Prepositional and Indirect Objects}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{preposition}
\index{prepositional objects}
\subsection*{Vocabulary}
\index{\textit{lon}}
\index{\textit{kepeken}}
\index{\textit{tawa}}
\index{\textit{kama}}       
\index{\textit{kiwen}}
\index{\textit{kon}}
\index{\textit{pana}}
\index{\textit{poki}}
\index{\textit{toki}}
\index{exist}
\index{on}
\index{at}
\index{in}
\index{be!in}
\index{be!at}
\index{be!on}
\index{with}
\index{using}
\index{go}
\index{move}
\index{for}
\index{come}
\index{happen}
\index{cause}
\index{stone}
\index{rock}
\index{air}
\index{atmosphere}
\index{spirit}
\index{wind}
\index{square}
\index{block}
\index{stairs}
\index{give}
\index{send}
\index{release}
\index{emit}
\index{container}
\index{bowl}
\index{glass}
\index{cup}
\index{box}
\index{language}
\index{talk}
\index{speak}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vocabulary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supertabular}{p{2,5cm}|ll}
\textbf{\dots lon} && \textit{adjective}: true, existing, correct, real, genuine \\ % no-dictionary
\textbf{lon} && \textit{noun}: existence, being, presence \\ % no-dictionary
\textbf{\dots , lon \dots} && \textit{preposition}: be (located) in/at/on \\ % no-dictionary
\textbf{lon} && \textit{verb intransitive}: to be there, to be present, to be real/true, to exist \\ % no-dictionary
 && \\ % no-dictionary
\textbf{kepeken} && \textit{noun}: use, usage, tool \\ % no-dictionary
\textbf{\dots , kepeken \dots} && \textit{preposition}: with \\ % no-dictionary
\textbf{kepeken} && \textit{verb intransitive}: to use \\ % no-dictionary
\textbf{kepeken \dots} && \textit{auxiliary verb}: to use \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots tawa} && \textit{adjective}: moving, mobile \\ % no-dictionary
\textbf{\dots tawa} && \textit{adverb}: moving, mobile \\ % no-dictionary
\textbf{tawa} && \textit{noun}: movement, transportation \\ % no-dictionary
\textbf{\dots , tawa \dots} && \textit{preposition}: to, in order to, towards, for, until \\ % no-dictionary
\textbf{tawa} && \textit{verb intransitive}: go to, walk, travel, move, leave \\ % no-dictionary
\textbf{tawa (e \dots)} && \textit{verb transitive}: to move, to displace \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots kiwen} && \textit{adjective}: hard, solid, stone-like, made of stone or metal \\ % no-dictionary
\textbf{\dots kiwen} && \textit{adverb}: hard, solid, stone-like, made of stone or metal \\ % no-dictionary
\textbf{kiwen} && \textit{noun}: hard thing, rock, stone, metal, mineral, clay \\ % no-dictionary
\textbf{kiwen (e \dots)} && \textit{verb transitive}: to solidify, to harden, to petrify, to fossilize \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots kon} && \textit{adjective}: air-like, ethereal, gaseous \\ % no-dictionary
\textbf{\dots kon} && \textit{adverb}: air-like, ethereal, gaseous \\ % no-dictionary
\textbf{kon} && \textit{noun}: air, wind, smell, soul \\ % no-dictionary
\textbf{kon} && \textit{verb intransitive:}: to breathe \\ % no-dictionary
\textbf{kon (e \dots)} && \textit{verb transitive}: to blow away something, to puff away something \\ % no-dictionary
 && \\ % no-dictionary
\textbf{\dots pana} && \textit{adjective}: generous \\ % no-dictionary
\textbf{pana} && \textit{noun}: giving, transfer, exchange \\ % no-dictionary
\textbf{pana (e \dots)} && \textit{verb transitive}: to give, to put, to send, to place, to release, to emit, to cause \\ % no-dictionary
 && \\ % no-dictionary
\textbf{poki} && \textit{noun}: container, box, bowl, cup, glass \\ % no-dictionary
\textbf{poki (e \dots)} && \textit{verb transitive}: to box up, to put in, to can, to bottle \\ % no-dictionary
\end{supertabular} \\
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\subsection*{Prepositions}
\index{preposition}
\index{Object!prepositional}
\index{prepositional object}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
A preposition describes a relationship between other words in a sentence and stand in front of nouns or pronouns. 
In Toki Pona prepositions start a prepositional object. 
A slot for prepositions is therefore only at the beginning of a prepositional object. 
A prepositional object is at the end of a block. 
Possible direct objects and intransitive objects (see below) are always in front of a prepositional object.
Like the other object types, a prepositional object is an optional part of a predicate phrase. 
It is recommended that you put a comma before a preposition.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Transitive Verbs vs. Intransitive Verbs}
\index{verb!intransitive}
\index{verb!transitive}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We've already learned transitive verbs. 
A transitive verb does something to the direct object. 
The direct object can be found out with' Who' or' What'. 
In front of the direct object there is a \textit{e}.

No direct object can follow an intransitive verb. 
Either no object or an intransitive object follows. 
The intransitive verb does not change the following object. 
You can't ask an intransitive object with' who' or' what'.
In front of the intransitive object there is no \textit{e}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{preposition!lon}
\index{verb!\textit{lon}}
\index{\textit{lon}!preposition}
\index{\textit{lon}!verb}
\subsection*{\textit{lon}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{lon} can be used as both a verb and a preposition. 
In this examples, \textit{lon} is used as a preposition.

\begin{supertabular}{p{5,5cm}|ll}
mi moku, lon tomo. && I eat in the house. \\
mi telo e mi, lon tomo telo. && I bathe myself in the restroom. \\
\end{supertabular} 

In this examples, \textit{lon} is used as an intransitive verb. 
Since there is no other predicate before \textit{lon} there must be a verb \textit{lon}.

\begin{supertabular}{p{5,5cm}|ll}
suno li lon sewi. && The sun is in the sky. \\
kili li lon poki. && The fruit is in the basket. \\
mi lon tomo. && I'm in the house. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{\textit{lon}!\textit{wile}}
\index{\textit{wile}!\textit{lon}}
\subsubsection*{Using \textit{lon} with \textit{wile}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Because \textit{lon} can be used as either a preposition or a intrantitive verb, the meaning of the sentence can be a bit confusing when used with \textit{wile}. 

\begin{supertabular}{p{5,5cm}|ll}
mi wile lon tomo. && I want to be at home. / I want in a house. \\
\end{supertabular} 

\index{trick!\textit{e ni}}
\index{\textit{e}!\textit{ni}}
\index{\textit{ni}!\textit{e}}
The sentence has at least two possible translations. 
The first translation states that the speaker wishes he were at home. 
The second translation states that the speaker wants to do something in a house. 
It's best to split this sentence up by a colon.
We can use a comma to force \textit{lon} as a preposition.

\begin{supertabular}{p{5,5cm}|ll}
mi wile, lon tomo. && I want to be at home. \\
\end{supertabular}

However, it is better to divide the sentence with a colon into two sentences.

\begin{supertabular}{p{5,5cm}|ll}
mi wile e ni: mi lon tomo. && I want this: I'm at home. \\
\end{supertabular} 

Toki Pona often uses this \textit{e ni:} trick. 
Before and after the colon has to be complete sentences. 
Toki Pona has no nested subordinate clauses.

\begin{supertabular}{p{5,5cm}|ll}
sina toki e ni tawa mi: sina moku. && You told me that you are eating. \\
\end{supertabular} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
\index{preposition!\textit{kepeken}}
\index{verb!\textit{kepeken}}
\index{\textit{kepeken}!preposition}
\index{\textit{kepeken}!verb}
\subsection*{\textit{kepeken}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Using \textit{kepeken} as an intransitive verb}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
\begin{supertabular}{p{5,5cm}|ll}
mi kepeken ilo. && I'm using tools. \\
sina wile kepeken ilo. && You have to use tools. \\
mi kepeken poki ni. && I'm using that cup. \\
\end{supertabular} 

In some other lessons \textit{kepken} is used as a transitive verb.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Using \textit{kepeken} as a preposition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{supertabular}{p{5,5cm}|ll}
mi moku, kepeken ilo moku. && I eat using a fork/spoon/ \\ && any type of eating utensil. \\
mi lukin, kepeken ilo suno. && I look using a flashlight.  \\
\end{supertabular} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{preposition!\textit{tawa}}
\index{verb!\textit{tawa}}
\index{\textit{tawa}!preposition}
\index{\textit{tawa}!verb}
\subsection*{\textit{tawa}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Using \textit{tawa} as a intransitive verb} \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Like \textit{lon}, \textit{tawa} as a intransitive verb doesn't have an \textit{e} after it. \\

\begin{supertabular}{p{5,5cm}|ll}
ona mute li tawa utala && They're going to the war. \\
sina wile tawa telo suli. && You want to go to the ocean. \\
ona li tawa sewi kiwen. && She's going up the rock. \\
\end{supertabular} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{\textit{tawa}!verb}
\index{verb!\textit{tawa}}
\subsubsection*{\textit{tawa} as a transitive verb}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{tawa} can also be a transitive verb.

\begin{supertabular}{p{5,5cm}|ll}
mi tawa e kiwen. && I'm moving the rock. \\
ona li tawa e len mi. && She moved my clothes. \\
\end{supertabular} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{\textit{tawa}!to}
\index{to!\textit{tawa}}
\subsubsection*{Using \textit{tawa} to mean 'to'}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

\begin{supertabular}{p{5,5cm}|ll}
mi toki, tawa sina. && I talk to you. \\
ona li lawa e jan, tawa ma pona. && He led people to the good land. \\
ona li kama, tawa ma mi. && He's coming to my country. \\
\end{supertabular} 

In this sentence the first \textit{tawa} is an intransitive verb.
The second \textit{tawa} is a preposition and introduces the prepositional object. 

\begin{supertabular}{p{5,5cm}|ll}
mi tawa, tawa tomo mi. && I'm going to my house. \\
\end{supertabular} 

\index{\textit{li}!\textit{li pona tawa mi}}
\index{\textit{li}!\textit{li ike tawa mi}}
\index{like!I like}
\index{I like}
In Toki Pona, to say that you (don't) like something, we have pattern, and the pattern uses \textit{tawa} as a preposition.

\begin{supertabular}{p{5,5cm}|ll}
ni li ' pona, tawa mi. && That is good to me. / I like that. \\
ni li ' ike, tawa mi && That is bad to me. / I don't like that. \\
\end{supertabular} 

\begin{supertabular}{p{5,5cm}|ll}
kili li ' pona, tawa mi. && I like fruit. \\
toki li ' pona, tawa mi. && I like talking. / I like languages. \\
utala li ' ike, tawa mi. && I don't like wars. \\
telo suli li ' ike, tawa mi. && I don't like the ocean. \\
\end{supertabular} 

\index{clauses}
Toki Pona does not use clauses. 
So for example, if you wanted to say 'I like watching the countryside,' it's best to split this into two sentences.

\begin{supertabular}{p{5,5cm}|ll}
mi lukin e ma. ni li ' pona, tawa mi. && I'm watching the countryside. This is good to me.\\
\end{supertabular} 

Of course, you could choose to say this same sentence using other techniques. 

\begin{supertabular}{p{5,5cm}|ll}
ma li pona lukin. && The countryside is good to look at. \\
\end{supertabular} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{\textit{tawa}!for}
\subsubsection*{Using \textit{tawa} to mean 'for'}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Okay, so \textit{tawa} essentially means 'to go' to or simply 'to', right? 
Not exactly. 
It can also mean 'for', as in this sentence.
 
\begin{supertabular}{p{5,5cm}|ll}
mi pona e tomo, tawa jan pakala. && I fixed the house for the disabled man. \\
\end{supertabular} 

Unfortunately, the trick of letting \textit{tawa} mean both 'to' and 'for' isn't without its drawbacks. 
Keep reading to see why. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\index{\textit{tawa}!adjective}
\index{adjective!\textit{tawa}}
\index{car}
\index{boat}
\index{ship}
\index{airplane}
\index{helicopter}
\subsubsection*{Using \textit{tawa} as an adjective}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\textit{tawa} is used as an adjective to make the phrase we use for 'car', 'boat' or 'airplane'.

\begin{supertabular}{p{5,5cm}|ll}
tomo tawa && car (moving construction) \\
tomo tawa telo && boat, ship \\
tomo tawa kon && airplane, helicopter \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Ambiguity of \textit{tawa}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
I want you to think about the following sentence for a few moments before continuing on. 
See if you can think of different meanings that it might have. 

\begin{supertabular}{p{5,5cm}|ll}
mi pana e tomo tawa sina. && ? \\   % no-dictionary
\end{supertabular} 

If \textit{tawa} is used as an adjective, then this sentence says 'I gave your car.' 
If it is used as a preposition, though, it could mean, 'I gave the house to you.' 
% So, how do you tell the difference? 
% You don't! (Insert evil, mocking laugh here.) 
You can put a comma before \textit{tawa}. 
Better is to write it in a different way.

\begin{supertabular}{p{5,5cm}|ll}
mi jo e tomo tawa sina. mi pana e ni tawa sina. && I have your car. I give it to you. \\
ni li tomo. mi pana e ni tawa sina. && This is a house. I give it to you. \\
\end{supertabular} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newpage
\index{\textit{kama}}
\index{\textit{kama}!\textit{tawa}}
\index{\textit{tawa}!\textit{kama}}
\subsection*{\textit{kama}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{\textit{kama} as a intransitive verb}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We've already briefly touched on this word, but it is quite common, and so we need to look at it a little more closely. 
First, it's used with the preposition \textit{tawa}.

\begin{supertabular}{p{5,5cm}|ll}
ona li kama, tawa tomo mi. && He came to my house. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{\textit{kama} as a transitive verb}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\index{cause}
\index{get}
\index{\textit{kama}!verb}
\index{verb!\textit{kama}}
It can also be used as a transitive verb. 
It means 'to cause' or 'to bring about'.

\begin{supertabular}{p{5,5cm}|ll}
mi kama e pakala. && I caused an accident. \\
sina kama e ni: mi wile moku. && You caused this: I want to eat. \\ && You made me hungry. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{\textit{kama} as Auxiliary Verb}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
It can also be used as auxiliary verb. 
One of the most common infinitives that it is used with is \textit{jo}, so we'll just cover it for now. 

\begin{supertabular}{p{5,5cm}|ll}
kama jo && get \\
mi kama jo e telo. && I got the water. \\
\end{supertabular} 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\subsection*{Practice (Answers: Page~\pageref{'prepositions01'})}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Please write down your answers and check them afterwards. 

Try to translate these sentences. 
You can use the tool \textit{Toki Pona Parser} (\cite{www:rowa:02}) for spelling and grammar check. 

\begin{supertabular}{p{5,5cm}|ll}
I fixed the flashlight using a small tool.  &&  \\ % no-dictionary
I like Toki Pona.   &&   \\ % no-dictionary
We gave them food. * &&    \\ % no-dictionary
This is for my friend.   &&   \\ % no-dictionary
The tools are in the container.   &&   \\ % no-dictionary
That bottle is in the dirt.  &&    \\ % no-dictionary
I want to go to his house using my car.  &&   \\  % no-dictionary
They are arguing.   &&   \\ % no-dictionary
&& \\ % no-dictionary
sina wile kama tawa tomo toki.   &&   \\ % no-dictionary
jan li toki, kepeken toki pona lon tomo toki.  &&    \\ % no-dictionary
mi tawa tomo toki. ona li ' pona tawa mi.  &&   \\  % no-dictionary
sina kama jo e jan pona, lon ni. **  &&   \\ % no-dictionary
\end{supertabular} 

\index{here}
\index{there}
* If you got that one wrong, think of the sentence like this: 
'We gave food to them.' It means the same thing. \\
** \textit{lon ni} means either 'here' or 'there'. 
Can you figure out what it literally means? 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% eof
