%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tokiponization Guidelines}
%
\label{'phonet_trans'}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
To create the Toki Pona version of a foreign name, you may use the following guidelines.
Also see Alphabet and sounds (Page~\pageref{'pronunciation_alphabet'}) for rules on what Toki Pona syllables and words are possible.
You can find a tool for transliterate of names in tokipona.net \cite{www:tokipona.net:01}.

\begin{itemize}
\item 
It is always better to translate the "idea" of a foreign word before attempting to create a new phonetic transcription that may not be recognizable by everyone. 
(Example: Jean Chrétien, Prime Minister of Canada = jan lawa pi ma Kanata, rather than jan Kesijen)
 \item 
Use the native pronunciation as a basis, rather than the spelling.
 \item 
If more than one language is spoken locally, use the dominant one.
 \item 
If it does not belong to any one language, use an international form. 
(Example: Atlantik = Alansi)
 \item 
Use the colloquial pronunciation that locals actually and commonly use, rather than the "proper" or standard pronunciation. 
(Example: Toronto = Towano, not Tolonto)
 \item 
If a person chooses to have a Toki Pona name, he can choose whatever he wants and does not necessarily have to follow these guidelines.
 \item 
Names of nations, languages, religions have already been established. 
If one is missing from the official list, make a suggestion on the Toki Pona discussion list.
 \item 
If possible, find a common root between the name of the nation, the language and the people. 
(Example: England, English, English(wo)man = Inli)
 \item 
Cities and locations can be given a Toki Pona name, but they will only have an official name if they are internationally known.
 \item 
If full Tokiponization would compromise intelligibility, you can always leave a foreign name as is.
\end{itemize} 

\subsection*{Syllables of Unoffial Words}

\begin{itemize}
 \item
Every syllable consists of a consonant and a vovel, plus an optional \textit{n}.
 \item
The next syllable after a optional \textit{n} cannot start with a \textit{n}.
 \item
The first syllable of a word does not need to beginn with a consonant.
 \item
The syllables \textit{ti} and \textit{tin} become \textit{si} and \textit{sin}.
 \item
The consonant \textit{w} cannot appear before \textit{o} or \textit{u}.
 \item
The consonant \textit{j} cannot appear before \textit{i}.
\end{itemize} 

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\subsection*{Phonetic Guidelines}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{itemize}
 \item 
Voiced plosives become voiceless. (Example: b = p, d = t, g = k)
 \item 
v becomes w.
 \item 
f becomes p.
 \item 
The trilled or tapped [r] of most world languages becomes l.
 \item 
The approximant r of languages like English becomes w.
 \item 
Any uvular or velar consonant becomes k, including the French or German r.
 \item 
At the end of a word, The sh sound may be converted to si. 
(Example: Lush = Lusi)
 \item 
The schwa can become any vowel in Toki Pona and is often influenced by neighbouring vowels for cute reduplication.
 \item 
It is better to keep the same number of syllables and drop a consonant than add a new vowel. 
(Example: Chuck = Sa, not Saku)
 \item 
When dealing with consonant clusters, the dominant plosive is generally kept, dropping fricatives such as [s] and laterals such al [l] first. 
(Example: Esperanto = Epelanto) 
You may also choose to keep the consonant at the head of the new syllable (Example: Atling = Alin).
 \item 
Approximants like [j] and [w] in consonant clusters can either be converted into a syllable of their own 
(Swe = Suwe; Pju = Piju) or dropped entirely (Swe = Se; Pju = Pu).
 \item 
In some cases, it is better to change the letter order slightly, rather than dropping a sound. 
(Ex: Lubnan = Lunpan, not Lupan or Lunan)
 \item 
Dental fricatives such as English th can either convert to t or s.
 \item 
The illegal syllables ti, wo and wu convert to si, o and u. 
(Example: Antarktika = Antasika)
 \item 
Affricates generally convert to fricatives. 
(Example: John = San, not Tan)
 \item 
Any nasal consonant at the end of a syllable converts to n. 
(Example: Fam = Pan)
 \item 
Nasal vowels (in French and Portuguese) also convert to syllable-final n.
 \item 
If necessary to preserve proper syllable structure, the consonant w or j can be inserted as a euphonic glide. (Example: Tai = Tawi; Nihon = Nijon; Eom = Ejon) It may also be possible to relocate a consonant that would have otherwise been dropped in the conversion. (Example: Monkeal = Monkela, not Monkeja; Euska = Esuka)
 \item 
Voiceless lateral consonants convert to s.
 \item 
If necessary, you may want to tweak a word to avoid a potentially misleading homonym. 
(Example: Allah = jan sewi Ila, not jan sewi Ala no God). 
If possible, use a related word in the source language rather than introducing an arbitrary change. 
(In Arabic, Allah actually means the God, whereas Illah means God.)
\end{itemize} 
% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% eof
